---
output:
  pdf_document:
    extra_dependencies:
    - changepage
    - enumitem
    - changepage
    - bbm
  html_document:
    df_print: paged
---
\begin{Large}\textbf{Carey Expected Return R Project Test}\end{Large}

## Exercise 1

#### a)

We first write a \texttt{R} function that determines the value of the parameters $\mu$ and $\sigma^{2}$ from the expectation $\mathbb{E}\{X\}$ and the variance $\mathbb{V}\{X\}$ of a univariate lognormal random variable $X\sim LogN(\mu,\sigma^{2})$:

```{r}
ln_params <- function(exp, var) {
  sig_sq <- log(var / exp^2 + 1)
  mu <- log(exp) - sig_sq / 2
  params <- list("mu" = mu, "sig_sq" = sig_sq)
  return(params)
}
```

Note: We managed to derive these formulas by rearranging the expressions for the mean and variance of a log-normal r.v.

#### b)

Next, we use the function created in a) to determine the parameters $\mu$ and $\sigma^{2}$ such that $\mathbb{E}\{X\}=3$ and $\mathbb{V}\{X\}=5$:

```{r}
params <- ln_params(3, 5)
ln_mu <- params$mu; ln_sig_sq <- params$sig_sq
ln_mu; ln_sig_sq
```

Next, we generate a large sample (of size $n=10,000$) from the distribution of $X$ with such parameters:

```{r}
n <- 10000
ln_sig <- sqrt(ln_sig_sq)
large_samp <- rlnorm(n, ln_mu, ln_sig)
```

After that, we plot the sample as dots:

```{r}
plot(large_samp, main = "Lognormal sample plot", xlab = "Sample indexes", ylab = "x values")
legend("topright", legend = "large_samp outcomes", pch = 1)
```

Next, we plot the histogram and overlay the the exact pdf for comparison:

```{r}
hist(large_samp, 50, FALSE, main = "Lognormal sample histogram w/ blue exact pdf overlayed", xlab = "x values")
curve(dlnorm(x, meanlog = ln_mu, sdlog = ln_sig), add = TRUE, col = "blue")
legend("topright", legend = "exact pdf", col = "blue", pch = "|")
```

Finally, we plot the empirical cdf and overlay the exact cdf for comparison:

```{r}
plot(ecdf(large_samp), main = "Red lognormal sample ecdf w/ blue exact cdf overlayed", xlab = "x values", ylab = "accumulated density", col = "red")
curve(plnorm(x, meanlog = ln_mu, sdlog = ln_sig), add = TRUE, col = "blue")
legend("bottomright", legend = c("ecdf", "exact cdf"), col = c("red", "blue"), pch = "|")
```

## Exercise 2

#### a)

First, we load in a library we will require for our simulation in b) and then proceed to write the function:

```{r}
# Load MASS library for use of mvrnorm function
library(MASS)

fit_locdisp_mlfp <- function(eps_t_mat, p_t_vec, nu, thresh) {
  # Extract length of vector and matrix, width of matrix
  t_bar <- length(p_t_vec)
  cols <- ncol(eps_t_mat)
  
  # 0. Initialize
  mu_vec <- vector("numeric", cols)
  for (t in c(1:t_bar)) {
    mu_vec <- mu_vec + p_t_vec[t] * t(eps_t_mat[t,])
  }
  sig_sq_hfp_mat <- matrix(rep(vector("numeric", cols), cols), cols, cols)
  for (t in c(1:t_bar)) {
    dif <- t(eps_t_mat[t,]) - mu_vec
    sig_sq_hfp_mat <- sig_sq_hfp_mat + p_t_vec[t] * t(dif) %*% dif
  }
  sig_sq_mat <- matrix(rep(vector("numeric", cols), cols), cols, cols)
  if (nu > 2) {
    sig_sq_mat <- ((nu - 2) / nu) * sig_sq_hfp_mat
  } else {
    sig_sq_mat <- sig_sq_hfp_mat
  }
  
  # Construct norm values to be updated after each run to check for convergence
  rel_euc_norm <- thresh * 2
  rel_fro_norm <- thresh * 2
  
  while ((rel_euc_norm > thresh) & (rel_fro_norm > thresh)) {
    # 1. Update weights and FP
    w_t_vec <- vector("numeric", t_bar)
    for (t in c(1:t_bar)) {
      dif <- t(eps_t_mat[t,]) - mu_vec
      w_t_vec[t] <- (nu + t_bar) / (nu + (dif %*% t(sig_sq_mat) %*% t(dif)))
    }
    sum_prod_p_w <- sum(p_t_vec * w_t_vec)
    q_t_vec <- vector("numeric", t_bar)
    for (t in c(1:t_bar)) {
      q_t_vec[t] <- p_t_vec[t] * w_t_vec[t] / sum_prod_p_w
    }
  
    # 2. Update output
    mu_vec_old <- mu_vec
    for (t in c(1:t_bar)) {
      mu_vec <- mu_vec + q_t_vec[t] * t(eps_t_mat[t,])
    }
    sig_sq_mat_old <- sig_sq_mat
    for (t in c(1:t_bar)) {
      dif <- t(eps_t_mat[t,]) - mu_vec
      sig_sq_hfp_mat <- sig_sq_hfp_mat + q_t_vec[t] * t(dif) %*% dif
    }
    
    # Update norm values to check for convergence; return to 1) if no convergence
    rel_euc_norm <- (norm(mu_vec - mu_vec_old)) / norm(mu_vec_old)
    rel_fro_norm <- (det(sig_sq_mat - sig_sq_mat_old)) / det(sig_sq_mat_old)
  }
  # 3. If convergence, output (mu_vec, sig_sq_vec)
  params <- list("mu_vec" = mu_vec, "sig_sq_mat" = sig_sq_mat)
  return(params)
}
```

#### b)

Next, we test the function:

```{r}
num_runs <- 1000
df <- 100
conv_thresh <- 10^(-9)
num_runs_inv <- 1 / num_runs
p_t_bvn_vec <- rep(num_runs_inv, num_runs)
mu_input <- c(0, 0)
Sigma_input <- cbind(c(1, 0), c(0, 1))
eps_t_bvn_mat <- mvrnorm(num_runs, mu_input, Sigma_input)
fit_locdisp_mlfp(eps_t_bvn_mat, p_t_bvn_vec, df, conv_thresh)
```

As these maximum likelihood estimates very closely resemble the vector of means and covariance matrix parameters we passed into the mvnorm function to perform our $1000$ simulations from the standard bivariate normal distribution, this test provides evidence of the functionality of our fit_locdisp_mlfp function.