---
output:
  pdf_document:
    extra_dependencies:
    - changepage
    - enumitem
    - changepage
    - bbm
  html_document:
    df_print: paged
---
\begin{Large}\textbf{Carey Expected Return R Project Test}\end{Large}

## Exercise 1

#### a)

We first write a \texttt{R} function that determines the value of the parameters $\mu$ and $\sigma^{2}$ from the expectation $\mathbb{E}\{X\}$ and the variance $\mathbb{V}\{X\}$ of a univariate lognormal random variable $X\sim LogN(\mu,\sigma^{2})$:

```{r}
find_params <- function(exp, var) {
  sig_sq <- log(var / exp^2 + 1)
  mu <- log(exp) - sig_sq / 2
  return(list("mu" = mu, "sig_sq" = sig_sq))
}
```

Note: We managed to derive these formulas by rearranging the expressions for the mean and variance of a log-normal r.v.

#### b)

Next, we use the function created in a) to determine the parameters $\mu$ and $\sigma^{2}$ such that $\mathbb{E}\{X\}=3$ and $\mathbb{V}\{X\}=5$:

```{r}
params <- find_params(3, 5)
ln_mu <- params$mu; ln_sig_sq <- params$sig_sq
ln_mu; ln_sig_sq
```

Next, we generate a large sample (of size $n=10,000$) from the distribution of $X$ with such parameters:

```{r}
n <- 10000
ln_sig <- sqrt(ln_sig_sq)
large_samp <- rlnorm(n, ln_mu, ln_sig)
```

After that, we plot the sample as dots:

```{r}
plot(large_samp, main = "Lognormal sample plot", xlab = "Sample indexes",
     ylab = "x values")
legend("topright", legend = "large_samp outcomes", pch = 1)
```

Next, we plot the histogram and overlay the the exact pdf for comparison:

```{r}
hist(large_samp, 50, FALSE,
     main = "Lognormal sample histogram w/ blue exact pdf overlayed",
     xlab = "x values")
curve(dlnorm(x, meanlog = ln_mu, sdlog = ln_sig), add = TRUE, col = "blue")
legend("topright", legend = "exact pdf", col = "blue", pch = "|")
```

Finally, we plot the empirical cdf and overlay the exact cdf for comparison:

```{r}
plot(ecdf(large_samp),
     main = "Red lognormal sample ecdf w/ blue exact cdf overlayed",
     xlab = "x values", ylab = "accumulated density", col = "red")
curve(plnorm(x, meanlog = ln_mu, sdlog = ln_sig), add = TRUE, col = "blue")
legend("bottomright", legend = c("ecdf", "exact cdf"), col = c("red", "blue"),
       pch = "|")
```

## Exercise 2

#### a)

First, we load in a library we will require for our simulation in b) and then proceed to write the function:

```{r}
# Load MASS library for use of mvrnorm function
library(MASS)

fit_locdisp_mlfp <- function(eps_mat, p_vec, nu, thresh) {
  # Extract length of p_vec (also width of eps_mat) and num_rows
  t_bar <- length(p_vec)
  num_rows <- nrow(eps_mat)
  
  # Introduce helper functions for constructing m_vec, s_mat
  construct_m_vec <- function(vec) {
    # simple matrix multiplication
    m_vec_ <- eps_mat %*% vec
    return(m_vec_)
  }
  construct_s_mat <- function(vec, m_vec_) {
    # Replace each 2*1 col of eps_mat w/ 2*2 prod
    mat_prod <- matrix(apply(eps_mat, 2, function(eps) (eps - m_vec_)
                             %*% t(eps - m_vec_)), 2)
    # Multiply each 2*2 square w/in matrix by corresponding vec val
    weighted_mat_prod <- sweep(mat_prod, 2, rep(vec, each = 2), '*')
    # Sum up 2*2 squares w/in matrix
    s_mat_ <- matrix(c(rowSums(weighted_mat_prod[, seq(1, 2*t_bar, 2)]),
                       rowSums(weighted_mat_prod[, seq(2, 2*t_bar, 2)])), 2)
  }
  # Learned sweep method from
  # https://intellipaat.com/community/24687/multiply-rows-of-matrix-by-vector
  
  # 0. Initialize
  m_vec <- construct_m_vec(p_vec)
  s_mat <- construct_s_mat(p_vec, m_vec)
  if (nu > 2) {
    s_mat <- ((nu - 2) / nu) * s_mat
  }
  
  # Construct norm values to be updated after each run to check for convergence
  rel_euc_norm <- thresh * 2
  rel_fro_norm <- thresh * 2
  
  while ((rel_euc_norm > thresh) || (rel_fro_norm > thresh)) {
    # 1. Update weights and FP
    w_vec <- (nu + num_rows) /
      (nu + apply(eps_mat, 2, function(eps) t(eps - m_vec) %*% t(s_mat) %*%
                    (eps - m_vec)))
    q_vec <- p_vec * w_vec / sum(p_vec * w_vec)
    
    # 2. Update output
    m_vec_old <- m_vec; m_vec_old
    m_vec <- construct_m_vec(q_vec)
    s_mat_old <- s_mat; s_mat_old
    s_mat <- construct_s_mat(q_vec, m_vec)
    
    # Update norm values to check for convergence; return to 1) if no
    # convergence
    rel_euc_norm <- (norm(m_vec - m_vec_old)) / norm(m_vec_old); rel_euc_norm
    rel_fro_norm <- (det(s_mat - s_mat_old)) / det(s_mat_old); rel_fro_norm
  }
  # 3. If convergence, output (mu_vec, sig_sq_vec)
  return(list("m_vec" = m_vec, "s_mat" = s_mat))
}
```

#### b)

Next, we test the function:

```{r}
n <- 1000
mu <- c(0, 0)
Sigma <- matrix(c(1, 0, 0, 1), 2)

eps_mat <- t(mvrnorm(n, mu, Sigma))
p_vec <- rep(1 / n, n)
nu <- 100
thresh <- 10^(-9)
fit_locdisp_mlfp(eps_mat, p_vec, nu, thresh)
```

As these maximum likelihood estimates very closely resemble the vector of means and covariance matrix parameters we passed into the mvnorm function to perform our $1000$ simulations from the standard bivariate normal distribution, this test provides evidence of the functionality of our fit_locdisp_mlfp function.